---
title: "Intro to Data Analysis with R: Exercise 1 Solutions"
subtitle: UCI Data Science Initiative
# author: "Chris Galbraith"
output: html_document
---

## Introduction

The first set of exercises will deal with reading a dataset into R, exploring various structural and content-related features of the data, and manipulating the dataset so that it is in a form we can use later for analyses. 

We will be using the Auto MPG Data Set, available on the UCI Machine Learning Repository; https://archive.ics.uci.edu/ml/datasets/Auto+MPG

The data consists of a collection of automobile records from 1970 to 1982 containing the following variables:

1. Miles per Gallon (mpg) 

2. Number of Cylinders

3. Engine Displacement (in cubic inches)

4. Horsepower

5. Weight (in pounds) 

6. Acceleration

7. Model Year

8. Origin: where the data originated from (ignore this)

9. Car Name

We will be focusing on the relationships between miles per gallon (mpg) and various other features of the car (such as model year, weight, number of cylinders, etc.). 


### Part A - Data Input
**A.1** Open a new R script to write and save your code for the exercises. Save this file in your local copy of the IDA-with-R-master directory, e.g., ```IDA-with-R-master/my_exercises/exercise_1.R```.

**A.2** Read in the Auto MPG data to a data frame named ```auto``` from the following url using ```read.table()```:  https://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data-original HINT: Run ```?read.table()``` and read about how to use a url as a file path.
```{r echo=TRUE}
auto <- read.table("https://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data-original")
```

**A.3** Rename the variables (columns) using the following conventions: "mpg", "cyl", "disp", "hp", "weight", "acc", "model.yr", "origin", "name". HINT: You will need to use the ```names``` attribute of the data frame (i.e., ```names(auto)```).
```{r echo=TRUE}
names(auto) <- c("mpg", "cyl", "disp", "hp", "weight", "acc", "model.yr", "origin", "name")
```

**A.4** Convert ```cyl``` into a factor variable using ```factor()```. Convert ```name``` into a character vector using ```as()```.
```{r echo=TRUE}
auto$cyl <- factor(auto$cyl)
auto$name <- as(auto$name, "character")
```

**A.5** Use the ```head()``` function to look at the first few rows of the data and make sure it looks like it was correctly loaded. You can compare the output here to the raw data by opening the url in A.2.
```{r echo=TRUE}
head(auto)
```


### Part B - Missing Values
**B.1** Locate the observations (rows) with missing data using ```is.na()```. HINT: You may want use ```which()``` with ```arr.ind=TRUE``` to return the (row, column) locations of the missing values.
```{r echo=TRUE}
missing <- which(is.na(auto), arr.ind = TRUE)
missing
```

**B.2** Look at the missing observations by subsetting the ```auto``` data frame.
```{r echo=TRUE}
auto[missing[,1], ]
```

**B.3** Which variables are missing? What are the implications of this missingness? 

*MPG is the response variable, we will need to predict these after fitting models.*

*Horsepower is a predictor variable, we  need to investigate its relationship with MPG.*


### Part C - Sorting
**C.1** Sort the Auto MPG data in descending order by mpg and store the result into a data frame named ```auto.sorted```. HINT: You will need to use ```order()``` with ```na.last=NA``` so that the values with missing mpg are not in the sorted data frame.
```{r echo=TRUE}
sort.index <- order(auto$mpg, decreasing = TRUE, na.last = NA)
auto.sorted <- auto[sort.index, ]
```

**C.2** Look at the observations with the top five values for mpg using ```head()```. 
```{r echo=TRUE}
head(auto.sorted, 5)
```

**C.3** Look at the observations with the bottom five values for mpg using ```tail()```.
```{r echo=TRUE}
tail(auto.sorted, 5)
```

**C.4** Do you notice any patterns with these two groups? HINT: You may need to do some Googling about these vehicles. 

*Answers will vary.*

### Part D - String Manipulation
**D.1** Locate the observations with diesel engines. HINT: If a vehicle has a diesel engine, it will mention "diesel" in the name of the car. Use the ```grep()``` function do accomplish this.
```{r echo=TRUE}
diesel.index <- grep("diesel", auto$name)
diesel.index
```

**D.2** Create a new variable (column) in the ```auto``` data frame called ```diesel``` such that ```auto$diesel = 1``` if the car has a diesel engine and ```0```, otherwise.
```{r echo=TRUE}
auto$diesel <- 0
auto$diesel[diesel.index] <- 1
```

**D.3** Coerce ```auto$diesel``` into a factor variable using ```as()```.
```{r echo=TRUE}
auto$diesel <- factor(auto$diesel)
```

**D.4** Look at the structure of the ```auto``` data frame using ```str()``` to make sure that this was done correctly.
```{r echo=TRUE}
str(auto)
```

**D.5** Save your data set as an R data (```.Rda```) file in the data directory (i.e.,  ```"IDA-with-R-master/data/auto_mpg_v2.Rda"```) using the ```save()``` function.
```{r echo=TRUE, eval=FALSE}
save(auto, file=here::here("data", "auto_mpg_v2.Rda"))
```


### Part A 
**A.1** Open a new R script and save it in the directory you created in Part A.1 of Exercise 1. Then, load the Auto MPG data set with the additional variable ```diesel``` using the ```load()``` function and file path from the end of Exercise 1. 
```{r echo= TRUE}
load(here::here("data", "auto_mpg_v2.Rda"))
```

**A.2** Using the ```summary()``` function, look at descriptive statistics for the Auto MPG data. What do you notice?
```{r echo= TRUE}
summary(auto)
```

### Part B
**B.1** Plot a relative frequency histogram of the response variable, MPG. Color the boxes grey. Make sure to name the plot and axes.
```{r echo=TRUE}
hist(auto$mpg, freq=FALSE, col = "grey", main = "Histogram of MPG", xlab = "MPG")
```

**B.2** Add a density curve to the histogram you plotted in B.1 using the ```lines()``` and ```density()``` functions. Color it red using the ```col``` argument. HINT: You need to change the value of ```na.rm``` in the density function to avoid an error. You also need to make sure that the histogram is on the proper scale!
```{r echo=TRUE}
hist(auto$mpg, freq=FALSE, col = "grey", main = "Histogram of MPG", xlab = "MPG")
lines(density(auto$mpg, na.rm=TRUE), col="red")
```

**B.3** Add a vertical line to the plot from B.2 at the median of MPG. Make it dashed using the ```lty``` arguement, and make it red. HINT: You may need to use ```na.rm``` again.
```{r echo=TRUE}
hist(auto$mpg, freq=FALSE, col = "grey", main = "Histogram of MPG", xlab = "MPG")
lines(density(auto$mpg, na.rm=TRUE), col="red", main="Densities of Samp1 and Samp2", xlab="")
abline(v = median(auto$mpg, na.rm=TRUE), col = "red", lwd = 2, lty=2)
```


### Part C
Create a boxplot of MPG grouped by the number of cylinders. Color the boxes grey. Make sure to name the plot and axes.
```{r echo=TRUE}
boxplot(mpg ~ cyl, data = auto, col = "grey",
        main = "Distribution of MPG by Number of Cylinders",
        xlab = "Cylinders", ylab = "MPG")
```


### Part D
Create a scatterplot matrix of the Auto MPG data. Include the following variables: "mpg", "cyl", "disp", "hp", "weight", "acc", "model.yr". What relationships do you see? Is there any confounding? Which variables should be included in a regression? 
```{r echo=TRUE, warning=FALSE}
library(car)
scatterplotMatrix( auto[, c("mpg", "cyl", "disp", "hp", "weight", "acc", "model.yr")] )
```

