---
title: Packages
subtitle: CSULB Intro to R
date: April 13, 2018
#author: Chris Galbraith, Micah Jackson
output: slidy_presentation
smaller: yes
---

## Packages
+ Used to share R code
+ Composed of code, data, documentation, and tests
+ Allow for easy, transparent and cross-platform extension of the R base system
+ Over 12,000 packages available on the Comprehensive R Archive Network (CRAN)
    + [Full list available packages](https://cran.r-project.org/web/packages/) 
+ For more info on how R packages are developed, please read [Creating R Packages: A Tutorial](http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf) by Friedrich Leisch.


## Terminology
+ *Package*: an extension of the R base system with code, data and documentation in a standardized format
+ *Base packages*: part of the R source tree, maintained by R Core (what you downloaded before coming today)
+ *Repository*: a website that hosts a package for installation
+ *Source*: the original version of a package with human-readable text and code
+ *Library*: a directory containing installed packages (this is on your machine)


## Installing Packages
There are three main ways to install a package in R:

1. Installing from CRAN: install a package directly from the repository
    + Using R studio: ```Tools > Install Packages...```
    + From R console: ```install.packages("package name")```
2. Installing from source: first download the add-on R package and then type the following in your console:
    + ```install.packages("path_to_file", repos = NULL, type = "source")``` 
3. Installing a variety of sources using the ```devtools``` package (must have it installed first!)
    + Github:  ```devtools::install_github("repo name")```
    + URL:  ```devtools::install_url("url")```
    + Other methods available, too
    

## Using Packages
There are two ways to access code and data from an installed package:

1. Loading the package to the current R workspace with the ```library()``` function
    + ```library(package)``` loads all contents of ```package``` and makes them accessible until R restarts
    + This is the recommended way to use packages

2. Directly referencing the code/data with the ```::``` operator
    + Instead of loading an entire package, we can instead temporarily access its contents 
    + ```package::name``` accesses ```name``` from ```package``` for a one-time use
    + Useful when you only need to use a single object once or twice 


## Install some packages from CRAN
Let's install ```devtools, here, and car``` from CRAN using the console:
```{r echo=TRUE, eval=FALSE}
install.packages("devtools", dependencies = TRUE)
install.packages("here", dependencies = TRUE)
install.packages("car", dependencies = TRUE)
```


## Using a Package
Now, let's use ```devtools``` to install ```BRRR``` from Github (more info [here](https://github.com/brooke-watson/BRRR)). 

Method 1:
```{r echo=TRUE, eval=FALSE}
library(devtools)  # load the devtools package
install_github("brooke-watson/BRRR")  # use the function install_github() to install BRRR
```

Method 2: 
```{r echo=TRUE, eval=FALSE}
devtools::install_github("brooke-watson/BRRR")  # use the function install_github() from devtools to install BRRR
```

Play around with BRRR!


## Popular Packages
+ To visualize data:
    + ```ggplot2```: to create beautiful graphics
    + ```googleVis```: to use Google Chart tools
+ To report results:
    + ```shiny```: to create interactive web-based apps
    + ```knitr```: to combine R codes and Latex/Markdown codes
    + ```slidify```: to build HTML 5 slide shows
+ To write high-performance R code:
    + ```rcpp```: to write R functions that call C++ code
    + ```data.table```: to organize datasets for fast operations
    + ```parallel```: to use parallel processing in R


## Helpful Resources
+ [Khan Academy](https://www.khanacademy.org/): free courses on programming, math, etc.
+ [CRAN](https://cran.r-project.org/): R manuals, FAQs, the R Journal
+ [Rbloggers](https://www.r-bloggers.com/): blog posts written by other R users
+ [RStudio cheat sheets](https://www.rstudio.com/resources/cheatsheets/)
+ [Stackoverflow](https://stackoverflow.com/): online community for asking and answering questions on a variety of topics (useful for searching error messages or for help with problem strategies)
+ [R for Data Science](http://r4ds.had.co.nz/): free online book by creators of RStudio


## DAY 2 - April 27

+ **Exploratory Data Analysis**: all the necessary tools to investigate your data before performing any formal modeling--from summary statistics to visualization including plotting histograms, boxplots, and scatterplots
+ **Linear Regression**: everything you need to know to begin fitting linear models--from simple t-tests to estimation of regression coefficients, variable selection, model diagnostics, and prediction
+ **Logistic Regression**: the basics of generalized linear models (GLMs) with an emphasis on binary response data--we extend the theory and modeling strategies of linear regression