---
title: "Exercise 4 Solutions"
subtitle: CSULB Intro to R
# author: Chris Galbraith, Micah Jackson
date: April 13, 2018
output: html_document
---


## Introduction
In this exercise, we will practice modifying and writing functions.


## Part A
1. Look again at your ```findTallest()``` function. How would we use this to find the age of the oldest person in the room? Write a function ```findOldest()``` that does performs this task.

You can actually use the ```findTallest()``` function without modification. Just enter in a vector of ages instead of heights. However, one way of rewriting your function may look like
```{r, evaluate=FALSE }
findOldest <- function(A){
  # initialize oldest
  oldest <- A[1]
  
  # search through all ages in A
  for(age in A){
    if(age > oldest){oldest <- age}
  }
  
  # spit out the oldest age after for-loop has run
  return(oldest)
}
```

2. Modify your ```findTallest()``` function to find the shortest person in the room. Name this new function ```findShortest()```.

Here, we just need to flip the ```>``` to ```<```.
```{r, evaluate=FALSE }
findShortest <- function(H){
  # initialize shortest
  shortest <- H[1]
  
  # search through all heights in H
  for(height in H){
    if(height < shortest){shortest <- height}
  }
  
  # spit out the shortest height after for-loop has run
  return(shortest)
}
```

3. Use one of your functions to find the tallest or shortest height from the vector ```theirHeights <- c(63, 58, NA, 72, 65)```. What happens? Modify your function to handle this missing data. HINT: There are several methods for this. If you wish to use an if/else statement, consider the function ```is.na()``` (```?is.na```). Alternatively, google "handling missing values R" and see what solutions you can find.

Solution 1: for-loop NAs to 0
```{r, evaluate=FALSE }
theirHeights <- c(63, 58, NA, 72, 65)

findTallest <- function(H){
    
    # for-loop to turn NAs to 0s
    for(i in 1:length(H)){
        if(is.na(H[i])){
            H[i] <- 0
        }
    }
    
   # initialize tallest
   tallest <- H[1]
  
   # search through all heights in H
   for(height in H){
     if(height > tallest){tallest <- height}
   }
  
   # spit out the tallest height after for-loop has run
   return(tallest)
}
```


Solution 2: Logical indexing NAs to 0
```{r, evaluate=FALSE }
findTallest <- function(H){

   # vectorized version of turning NAs to 0s
   H[is.na(H)] <- 0
    
   # initialize tallest
   tallest <- H[1]
  
   # search through all heights in H
   for(height in H){
     if(height > tallest){tallest <- height}
   }
  
   # spit out the tallest height after for-loop has run
   return(tallest)
}
```


Solution 3: Omit NAs
```{r, evaluate=FALSE }
findTallest <- function(H){
   
   # remove the NAs from H
   H <- na.omit(H)
       
   # initialize tallest
   tallest <- H[1]
  
   # search through all heights in H
   for(height in H){
     if(height > tallest){tallest <- height}
   }
  
   # spit out the tallest height after for-loop has run
   return(tallest)
}
```


## Part B
Write a function that will take in 4 numbers and return them in order from least to greatest. 
HINT:
```{r eval=FALSE}
myNums <- c(9,3,27,6)

mySort <- function(numbers){

    # need to repeat at most the number of items you're sorting
    for(i in 1:length(numbers)){
        # compare all of your numbers. don't need to go to length(numbers)
        # because it has already been compared in the previous iteration.
        for(j in 1:(length(numbers)-1)){
            if(numbers[j] > numbers[j+1]){
                # reassign the numbers in correct order
                numbers[c(j,j+1)] <- swap(numbers[j], numbers[j+1])
            }
        }
    }
    # return your sorted numbers
    return(numbers)
}


mySort(c(9,3,27,6))
```


