---
title: Data Structures
subtitle: CSULB Intro to R
date: April 13, 2018
#author: Chris Galbraith, Micah Jackson
output: slidy_presentation
smaller: yes
---


## Agenda
1. Vectors and Matrices
2. Lists and Data Frames
3. Type Coercion
4. Troubleshooting


## Data Types in R
<!-- hierarchy? -->
+ R has 5 main atomic data types:
    + Logical
    + Integer
    + Numeric
    + Complex
    + Character


## Data Structures in R
1. One-dimensional:
    + Vectors
    + Lists
2. Multi-dimensional:
    + Matrices
    + Data frames

&nbsp;

+ Everything in R is an object
+ Objects can have attributes
    + e.g., names, dimension, length


## Vectors in R

+ A vector is the most basic structure in R
+ It is one-dimensional; its single dimension is its length
+ A vector of length *n* has *n* cells
+ Each cell can hold a single value, like a numeric or string value
    + In general, vectors can only hold ONE type of data 
  
```{r echo=TRUE}
numVec <- c(2,3,4)      # <- is the assigning operator
numVec
```


## Examples of Vectors
Examples of character, logical, and complex vectors:

```{r echo=TRUE}
intVec <- c(2L, 3L, 4L)
intVec

charVec <- c("red", "green", "blue")
charVec

logVec <- c(TRUE, FALSE, FALSE, T, F)
logVec

```


## Matrices
+ A matrix is a special case of a vector
    + Unlike vectors, matrices have a dimension attribute
    
```{r echo=TRUE}
myMat <- matrix(nrow = 2, ncol = 4)
myMat
attributes(myMat)
```


## Matrices

```{r echo=TRUE}
myMat <- matrix(1:8, nrow = 2, ncol = 4)
myMat # matrices are filled in column-wise
```


## A matrix is a special case of a vector

```{r echo=TRUE}
myVec <- 1:8
myVec
dim(myVec) <- c(2,4)
myVec
```
+ Similar to vectors, all elements of a matrix should be of the same data type
    + If not, R automatically coerces


## Other Ways to Create a Matrix

+ Intuitively, matrices seem to be a combination of vectors that are put next to each other (either column-wise or row-wise).

+ ```rbind()``` and ```cbind()``` (row bind and column bind) achieve this: 

```{r echo=TRUE}
vec1 <- 1:4
vec2 <- sample(1:100, 4, replace = FALSE)
vec3 <- sample(1:20, 4, replace=TRUE)
colMat <- cbind(vec1, vec2, vec3)
colMat
```


## Other Ways to Create a Matrix

```{r echo=TRUE}
vec1 <- 1:4
vec2 <- sample(1:100, 4, replace = FALSE)
vec3 <- rnorm(4, mean = 0, sd = 1)
rowMat <- rbind(vec1, vec2, vec3)
rowMat
```


## Factors

+ A factor is a vector used to specify a discrete classification (categorical values).
+ Factors can be ordered or un-ordered
+ Levels of a factor are better when labeled (self-descriptive)
    + Consider sex as (0, 1) as opposed to labeled ("F", "M")
    
```{r echo=TRUE}
Sex <- rep(c("Female", "Male"), times = 3)
Sex
SexFac1 <- factor(Sex)
SexFac1
```


## Factors

```{r echo=TRUE}
levels(SexFac1)
table(SexFac1)
```


## Factors

```{r echo=TRUE}
SexFac1 # levels are ordered alphabetically - 1st level = BaseLevel
SexFac2 <- factor(Sex, levels = c("Male", "Female"))
SexFac1
SexFac2
```


## Lists
+ Think of a list as a vector with the following main differences: 
    + Each element of a list can have its own data structure regardless of other elements
        + vector, matrix, another list
    + This means, each element can be of a different data type and a different length
    
```{r echo=TRUE}
myVec <- c(10, "R", 5L, T)
myVec
```


## Lists

```{r echo=TRUE}
myList <- list(10, "R", 5L, T)
myList
```
+ Elements of a list are shown with ```[[]]```
+ Elements of a vector are shown with ```[]```


## Data Frames
+ A data frame is a special list where all objects have equal length
+ A data frame looks very similar to a matrix; however, different columns in a data frame can be different data types

```{r echo=TRUE}
studentID <- paste("S#", sample(c(6473:7392), 10), sep = "")
score <- sample(c(0:100), 10)
sex <- sample(c("female", "male"), 10, replace = TRUE)
data <- data.frame(studentID = studentID, score = score, sex = sex)
head(data)
```


## Special Values
There are some special values in R:

  + Use ```L``` to refer to an integer value, e.g., ```1L```
  + R knows infinity: ```Inf```, ```-Inf```
  + ```NaN```: refers to "Not a number"
  
```{r echo=TRUE}
intVec <- c(1L, 2L, 3L, 4L) 
intVec

a <- Inf; b <- 0
rslt <- c(b/a, a/a)
rslt
```


## Missing Values
+ There are two kinds of missing values in R:
    + ```NaN```: stands for "Not a Number" and is a missing value produced by numerical computation.
    + ```NA```: stands for "Not Available" and is used when a value is missing
+ ```NaN``` is also considered as ```NA``` (the reverse is NOT true)
```{r echo=TRUE}
a <- c(1,2)
a[3]

b <- 0/0
b
```


## Data Type Coercion
+ In general, vectors CANNOT have mixed types of objects

```{r echo=TRUE, results='hide'}
numCharVec <- c(3.14, "a")
numCharVec                 # What do you expect to be printed? 

numLogVec <- c(pi, T)
numLogVec                   

charLogVec <- c("a", TRUE)
charLogVec 
```
+ The above are examples of implicit coercion 
+ Explicit coercion is also possible


## Data Type Coercion
+ ```as()```: explicitly coerces objects from one type to another

```{r echo=TRUE}
numVec <- seq(from = 1200, to = 1300, by = 15)
numVec
numToChar <- as(numVec, "character")
numToChar
numToChar==as.character(numVec)
logVec <- c(F, T, F, T, T)
as(logVec, "numeric")
as.numeric(logVec)
```



## Data Type Coercion
+ Coercion does not always work! Be careful about warnings:

```{r echo=TRUE, error=TRUE}

charVec <- c("2.5", "3", "2.8", "1.5", "zero")
as(charVec, "numeric")

charVec <- c("2.5", "3", "2.8", "1.5", zero)
```


## Summary
+ There are 5 atomic data types in R. Logical, integer, numeric, complex, character.
+ Vectors and matrices are used to store data of the same type.
+ Lists and data frames are used to store data of different types.
+ R con handle infinity (```Inf```), missing values (```NA```), and division by 0 (```NaN```)
+ Different data types inside of a vector or matrix will be coerced to the most flexible.


## Up Next
+ [Exercise 2](https://datumu.github.io/CSULB_Intro_R/exercises/exercise_2/ex_2.html)
+ Break





