---
title: "Exploratory Data Analysis"
subtitle: CSULB Intro to R
date: April 27, 2018
#author: "Chris Galbraith, Micah Jackson"
output: slidy_presentation
smaller: yes
---

```{r, include=FALSE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(vioplot)
```


## Agenda
1. Brief Review

2. Exploratory Data Analysis

    + Numeric Summary Statistics
    + Histograms
    + Boxplots
    + Scatterplot Matrices
    
    
## Vector and Matrix Review
+ Vectors and Matrices are data structures used to hold data of the same type
    + ```myVec <- c(1,2,3)```
    + ```myMatrix <- matrix(c('Ariel', 'Bill', 'Chris', 'Devon',```
    ```ncol=3, nrow=2)```
+ If the data is not of the same type, R will **coerce** the data to the most flexible type
+ We subset these structures using the ```[]``` operator
    + ```myVec[2]```, &nbsp; ```myVec[-3]```, &nbsp; ```myVec[myVec >= 2]```
    + ```myMatrix[2,1]```, &nbsp; ```myMatrix[3,]```, &nbsp; ```myMatrix[,1]```


## Data Frame Review
+ Data Frames are special lists in R that hold different data types
    + ```myDF <- data.frame(names=c('Albert', 'Bianca', 'Cat'),```
    ```height=c(71, 57, 64),```
    ```female=c(0,1,1))```
+ Each column must be the same length and must have the same data type within that column
+ We subset (named) data frames using the ```$``` operator. This can be combined with the ```[]``` operator.
    + ```myDF$names```
    + ```myDF$height[myDF$height > 60]```
    + ```myDF$height[myDF$female==1]```
    
    
## Function Review
+ Functions, ```function()```, are blocks of code or instructions that take some input and produce a desired output. 
+ The input is known as the arguments or parameters
+ In R, many functions have preset/default argument values so that you don't have to specify them each time. However, it's usually good practice to specify several arguments to make your output more readable.
    

## Reading Data
+ Before beginning a data analysis, we need to read in the appropriate data
+ This is generally handled with 1 of 2 functions: ```read.table()```, &nbsp; or ```read.csv()```
+ [Auto MPG data](https://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data)
    + ```autos <-```
    ```read.table(file='https://archive.ics.uci.edu/ml/```
    ```machine-learning-databases/auto-mpg/auto-mpg.data',```
    ```header=FALSE)```
+ [Abalone Whale Data](https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data)
    + ```whales <-```
    ```read.csv(file='https://archive.ics.uci.edu/ml/```
    ```machine-learning-databases/abalone/abalone.data',```
    ```header=FALSE)```
    
    
## Writing Data
+ Writing data is useful when you need to save a copy of your dataset to your local machine
+ This is usually done with ```write.table()``` and ```write.csv()```
    + ```write.table(object, filename, sep)```
    + ```write.csv(object, filename)```

    
## Some Useful Functions
+ ```str()```: a function to explain internal structure of an object
+ ```summary()```: a function that summarizes variables in a data frame 
    + Note: this function is also used to summarize results of model fitting functions, which we will go over in the afternoon.


## ```str()```
Compact way of understanding what an object is and what it contains

```{r, echo=TRUE}
str(mean)
str(matrix)
str(sample)
```


## ```str()```
After loading a data frame, it is often useful to use ```str()``` in order to understand the structure of your data. 

```{r echo=FALSE}
myDF <- data.frame(names=c('Albert', 'Bianca', 'Cat'), height=c(71, 57, 64), female=c(0,1,1))
```

```{r echo = TRUE, eval=TRUE}
str(myDF)
prestige <- read.csv('~/Documents/git/CSULB_Intro_R/data/prestige_v2.csv', header=TRUE)
# prestige <- read.table(file = here::here("data", "prestige_v2.csv"), 
#                        sep=",", 
#                        header=TRUE, 
#                        row.names=1)
str(prestige)
```


## ```summary()```
Another useful function for understanding your data by providing a numeric summary of each attribute (column).
```{r echo = TRUE, eval=TRUE}
summary(prestige)
```


## Exploratory Data Analysis
+ Before performing statistical analyses on your data, it is important to do exploratory data analysis (EDA) in order to better understand the variables and the relationships between them.

+ This can be done in many ways

    + Numeric summaries (e.g., using ```str()``` and ```summary()```)
    + Plots, plots, and more plots
    + More advanced methods (variograms, empirical covariance matrices, etc)
    
+ We will cover some of the basic plotting functions.


## Plotting in R
+ To explore the distribution of one variable:
    
    + 5-number summary
    + Boxplots
    + Histograms and Density Estimation
    + Violin plots
    

+ To explore relationships between variables:
    
    + Grouped Boxplots
    + Scatterplots
    + Scatterplot Matrices
    + Q-Q plot
    
    
## Useful plotting resources:

  + [Quick-R Graphical Parameters](https://www.statmethods.net/advgraphs/parameters.html)
  + [Colors in R](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf)
  + [Points and Line Types](http://www.cookbook-r.com/Graphs/Shapes_and_line_types/)
  + [```ggplot2```](http://ggplot2.tidyverse.org/)
  + Search for (or post) specific questions on [stackoverflow](https://stackoverflow.com/questions/tagged/r), which is a community that will answer questions & chose the best solutions via voting
  
+ *NOTE: This course provides a basic introduction to ```R```'s plotting capabilities. You can do much, much more elegant plots in ```R```!*


## 5 number summary
+ As seen in the ```summary()``` function, the distribution of a single numeric variable can be summed up wth 5 numbers. These are the minimum and max values, the $1^{st}$ and $3^{rd}$ quartiles (25% and 75%), and the median (50%).
+ R also throws in the mean to create a "6 number summary." We will focus on the 5 without the mean.

```{r, echo=TRUE}
summary(prestige$prestige)
```
<!-- explain 5 number summary and draw pics on board describing quantiles -->


## Box plots
+ A box plot, ```boxplot()```, is a handy graphical representation of the 5 number summary. 
+ Just as with the 5 number summary, the box plot tells us the min$^*$ and max$^*$ values, the $1^{st}$ and $3^{rd}$ quartiles, and the median. 
    + If outliers are present, by default defined as outside of 1.5$\times$*inter-quartile range*, they will be shown outside of the whiskers.

```{r, echo=TRUE}
boxplot(prestige$prestige, horizontal=TRUE, xlab='Prestige Scores')
```
<!-- build the plot adding one parameter at a time to explain parameters -->



## Histograms
+ Another useful graphic to visualize the distribution of a variable is a histogram, ```hist()```.
+ Reduced version of a "stem-and-leaf" plot
+ For each possible value of your variable (x-axis), a histogram displays the number of times that value appears in your data (y-axis). Counts can also be converted into proportions.
+ This is a natural plot for a discrete variable. For continuous variables, a histogram "bins" the observations into discrete units/intervals.
```{r echo = TRUE, eval=TRUE}
hist(prestige$prestige, freq = FALSE, 
     col = "grey",
     main = "Histogram of Prestige Score", 
     xlab = "Prestige Score")
```


## Histograms, contd.
+ Sometimes we would rather have a smoothed version of a histogram (i.e., a *density* function).
    + Not susceptible to number/location of bins like histograms are.
+ We can include this using the ```density()``` and ```lines()``` functions.
    + ```lines()```: takes coordinate pairs (in multiple input formats) and adds them to current figure connected by line segments
    + ```density()```: computes kernel density estimates (a smoothed histogram); see ```?density``` for more details
    
```{r echo = TRUE, eval=TRUE}
hist(prestige$prestige, freq=FALSE, col = "grey", main = "", xlab = "Prestige Score", ylim=c(0,0.022))
lines(density(prestige$prestige, bw='nrd0'), col = "red", lwd = 2)
```


## Violin Plots
+ So far we've seen how a distribution can be summarized by a 5 number summary, which can be visualized with a box plot. 
+ We've also seen how to view the shape of a distribution using histograms and density estimates.
+ `r emo::ji('thinking')` Can we combine these into one graphic?
    + Obviously. We made a whole slide for it.
+ Violin plots create boxplots, but overlay them with a mirrored density estimate

```{r echo=TRUE}
vioplot(prestige$prestige, horizontal=TRUE, col='dodgerblue2',
  names='Prestige Scores')
```
<!-- mention much better violin plotting capabilities (and more flexible in general) with ggplot -->


<!-- QUICK EXERCISE???? -->

## Exploring more than one variable
+ We've looked at ways to view the distribution of single variables, but how do we examine the relationship between 2 variables? 3? 42?
+ We'll focus on 2 for now.
+ Types of relationships:
    + Discrete - Discrete
    + Discrete - Continuous
    + Continuous - Continuous
    

## Discrete - Discrete
+ Many times, we'll be interested in a discrete response across classes of subjects, e.g. proportion of smokers to non-smokers across males/females.
+ This can usually be summed up with a table

|       | Male | Female |
|------:|:----:|:------:|
| Smoke |  16% |  12%   |
| Nah   |  41% |  31%   |

+ We could also overlay histograms, though this is more beneficial for ordered factors or continuous covariates.


## Overlay Histograms
+ For Discrete - Continuous types
+ When examining the distribution of a single variable, we can easily create a histogram using ```hist()```. We can extend this to creating a histogram of a single variable, but for multiple groups. 
+ Let's examine the ```prestige``` scores grouped by ```type```. 

```{r, echo=TRUE}
pp <- prestige$prestige
hist(pp[prestige$type=='bc'], breaks=10, main='Histograms of Prestige Scores by Job Type',
  xlab='Prestige Scores', col=adjustcolor('dodgerblue2',.4), xlim=c(0,100))
hist(pp[prestige$type=='wc'], breaks=10, col=adjustcolor('green', .4), add=TRUE)
hist(pp[prestige$type=='prof'], breaks=10, col=adjustcolor('purple', .4), add=TRUE)
```


## Adding Lines
+ While looking at these histograms (or any plot for that matter), we may wish to add lines indicating the location of the mean, median, quartiles, etc.
+ We can do this using the ```abline()``` function.
+ With ```abline()```, we can plot a vertical line, ```abline(v=...)```, horizontal line, ```abline(h=...)```, generic lines, ```abline(a=intercept, b=slope,...)```, and others.

```{r, echo=TRUE, eval=FALSE}
abline(v=mean(pp[prestige$type=='bc']), lwd=2, col='dodgerblue2')
abline(v=mean(pp[prestige$type=='wc']), lwd=2, col='green')
abline(v=mean(pp[prestige$type=='prof']), lwd=2, col='purple')
```

```{r, echo=FALSE, eval=TRUE}
hist(pp[prestige$type=='bc'], breaks=10, main='Histograms of Prestige Scores by Job Type',
  xlab='Prestige Scores', col=adjustcolor('dodgerblue2',.4), xlim=c(0,100))
hist(pp[prestige$type=='wc'], breaks=10, col=adjustcolor('green', .4), add=TRUE)
hist(pp[prestige$type=='prof'], breaks=10, col=adjustcolor('purple', .4), add=TRUE)

abline(v=mean(pp[prestige$type=='bc']), lwd=2, col='dodgerblue2')
abline(v=mean(pp[prestige$type=='wc']), lwd=2, col='green')
abline(v=mean(pp[prestige$type=='prof']), lwd=2, col='purple')
```

## Adding a Legend
+ With multiple features on one figure, a legend can help clearly convey what is plotted.
+ Use ```legend()``` to do this in R.
    + Very versatile function--look at its documentation 
    + Best to play around with this on your own, iterating through multiple plots until you get the legend to appear how you want it
    
```{r, echo=TRUE, eval=FALSE}
legend("topright", legend = c("BC", "WC", "PROF"),
       col = c("dodgerblue2", "green", "purple"), lty = 1, lwd = 2, bty = "n")
```

```{r echo=FALSE}
hist(pp[prestige$type=='bc'], breaks=10, main='Histograms of Prestige Scores by Job Type',
  xlab='Prestige Scores', col=adjustcolor('dodgerblue2',.4), xlim=c(0,100))
hist(pp[prestige$type=='wc'], breaks=10, col=adjustcolor('green', .4), add=TRUE)
hist(pp[prestige$type=='prof'], breaks=10, col=adjustcolor('purple', .4), add=TRUE)

abline(v=mean(pp[prestige$type=='bc']), lwd=2, col='dodgerblue2')
abline(v=mean(pp[prestige$type=='wc']), lwd=2, col='green')
abline(v=mean(pp[prestige$type=='prof']), lwd=2, col='purple')

legend("topright", legend = c("BC", "WC", "PROF"),
       col = c("dodgerblue2", "green", "purple"), lty = 1, lwd = 2, bty = "n")
```


## Grouped Violin plots
+ Used to investigate relationships between variables:
    + Continuous variable & factor (most common)
    + Two continuous variables (group one of them)
+ Let's look at violin plots of ```prestige``` grouped by ```type```
```{r echo = TRUE, eval=TRUE}
vioplot(pp[prestige$type=='bc'], pp[prestige$type=='wc'], pp[prestige$type=='prof'],
  horizontal=TRUE, names=c('BC', 'WC', 'PROF'))
```


## Scatterplots
+ We can also use scatterplots to visualize the relationship between two (continuous) variables.
+ We use the ```plot()``` function to do this
    + Extremely flexible--there are plot methods for a variety of objects! (e.g., plotting a regression object returns diagnostic plots... more on this later)
    + See the help documentation (```?plot```) and [Graphical Parameters](https://www.statmethods.net/advgraphs/parameters.html) for more details
+ For now, focus on the most simple plot where we specify the ```x``` and ```y``` coordinates.
    + ```x``` and ```y``` must be the same dimension
    
```{r echo = TRUE, eval=TRUE}
plot(x = prestige$education, y = prestige$prestige, pch = 20,
     main = "Prestige Score by Education",
     xlab = "Avg. Years of Education", ylab = "Prestige Score")
```


## Scatterplots, contd.
+ Let's overlay both a linear fit and smoother to our scatterplot. 
+ We can include this using the ```lowess()```, ```lines()```, and ```abline()``` functions.
    + Need to use ```lm()``` to fit a linear regression, more on this later
    + ```lowess()```: **LO**cally **WE**ighted **S**catterplot **S**moother; computes a smoothed fit using locally-weighted polynomial regression.
    
```{r echo = TRUE, eval=TRUE}
plot(prestige$education, prestige$prestige, pch = 20,
     main = "Prestige Score by Education",
     xlab = "Avg. Years of Education", ylab = "Prestige Score")

abline(reg = lm(prestige ~ education, data = prestige),
  col = "green", lwd = 2)  # linear regression

lines(lowess(x = prestige$education, y = prestige$prestige),
  col = "red", lwd = 2)  # smoother

legend("topleft", legend = c("Regression Line", "Smoother"),
  col = c("green", "red"), lwd = c(2,2), lty = 1, bty = "n")
```





## Scatterplot Matrices
+ The function ```scatterplotMatrix()``` (found in the ```car``` package we installed earlier) produces scatterplots between all variables in a data frame.
+ We can use direct ordering of the variables to control the order in which they are plotted.
```{r echo = TRUE, eval=TRUE, warning=FALSE}
library(car)
scatterplotMatrix( prestige[ ,c("prestige","education","income","women")] )
```


## Summary
+ We've seen how to explore single variables with boxplots, histograms and density estimatates, and a combination of these known as a violin plot.
+ For 2 variables, we created grouped violin plots, overlayed histograms, and scatterplots. 
+ Visualizations can be enhanced by adding lines, colors, labels, etc
+ There is always more than one way to view your data. Each perspective can give you a new understanding.


## Some common commands and arguments
* Plots
    + ```boxplot()```
        - horizontal=TRUE
    + ```hist()```
        - freq=TRUE
    + ```density()```
        - bw=c('nrd0', 'SJ', 5, 0.5,...)
    + ```vioplot::vioplot()```
        - horizontal=TRUE, names=c('name1', 'name2',...)
    + ```plot()```
        
* Common plot arguments (for **most** plots)
    + labeling: main, xlab, ylab
    + limits: xlim, ylim
    + add to existing plot: add=TRUE
    + point and line types and shapes: pch, lty, lwd
    + color: col
  
* Other useful functions
    + To add curves to existing plot: ```lines()```
    + To add straight line to existing plot: ```abline()```
    + To adjust color transparency: ```adjustcolor()```



## Next up
1. [Exercise 1](ADD ME)
2. Lunch

**Return at XXXX to discuss solutions to Exercise 2!**
