---
title: "Exploratory Data Analysis"
subtitle: CSULB Intro to R
date: April 27, 2018
#author: "Chris Galbraith, Micah Jackson"
output: slidy_presentation
smaller: yes
---

```{r, include=FALSE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Agenda
1. Brief Review

2. Exploratory Data Analysis

    + Numeric Summary Statistics
    + Histograms
    + Boxplots
    + Scatterplot Matrices
    
## Review
review day 1 materials here


## Some Useful Functions
+ ```str()```: a function to explain internal structure of an object
+ ```summary()```: a function that summarizes variables in a data frame 
    + Note: this function is also used to summarize results of model fitting functions, which we will go over in the afternoon.


## ```str()```
Compact way of understanding what an object is and what it contains
```{r echo=TRUE}
str(str)
str(sample)
```


## ```str()```
After loading a data frame, it is often useful to use ```str()``` in order to understand the structure of your data. 

```{r echo = TRUE, eval=TRUE}
prestige <- read.table(file = here::here("data", "prestige_v2.csv"), 
                       sep=",", 
                       header=TRUE, 
                       row.names=1)
str(prestige)
```


## ```summary()```
Another useful function for understanding your data by providing a numeric summary of each attribute (column).
```{r echo = TRUE, eval=TRUE}
summary(prestige)
```


## Exploratory Data Analysis
+ Before performing statistical analyses on your data, it is important to do exploratory data analysis (EDA) in order to better understand the variables and the relationships between them.

+ This can be done in many ways

    + Numeric summaries (e.g., using ```str()``` and ```summary()```)
    + Plots, plots, and more plots
    + More advanced methods (variograms, empirical covariance matrices, etc)
    
+ We will cover some of the basic plotting functions.


## Plotting in R
+ To explore the distribution of one variable:
    
    + Histograms
    + Boxplots

+ To explore relationships between variables:
    
    + Grouped Boxplots
    + Scatterplots
    + Scatterplot Matrices
    
+ Useful resources:

    + [Quick-R Graphical Parameters](https://www.statmethods.net/advgraphs/parameters.html)
    + [```ggplot2```](http://ggplot2.tidyverse.org/)
    + Search for (or post) specific questions on [stackoverflow](https://stackoverflow.com/questions/tagged/r), which is a community that will answer questions & chose the best solutions via voting
  
+ *NOTE: This course provides a basic introduction to ```R```'s plotting capabilities. You can do much, much more elegant plots in ```R```!*


## Histograms
+ One of the most basic plots in R is a histogram.
+ Let's plot a histogram of the response variable ```prestige```.
    + ```freq```: logical variable that controls the type of histogram (```TRUE``` gives counts, ```FALSE``` gives relative counts)
    + ```breaks```: controls the number of bins & bin locations; multiple ways to set this
    + We can add arguments to the function to change the bar colors, title, and axis labels.
```{r echo = TRUE, eval=TRUE}
hist(prestige$prestige, freq = FALSE, 
     col = "grey",
     main = "Histogram of Prestige Score", 
     xlab = "Prestige Score")
```


## Histograms, contd.
+ Let's add a dashed vertical line at the median of prestige using ```abline()```
    + Adds a straight line to the current figure
    + Can be specified multiple ways--intercept/slope, horizontal/vertical location, regression coefficients/object
    + Not unique to histograms--you can use ```abline``` on (almost) any figure
```{r echo = TRUE, eval=TRUE}
hist(prestige$prestige, freq = FALSE, col = "grey", 
     main = "Histogram of Prestige Score", xlab = "Prestige Score")
abline(v = median(prestige$prestige), col = "blue", lty = 2, lwd = 2)
```


## Histograms, contd.
+ Sometimes we would rather have a smoothed version of a histogram (i.e., a *density* function).
    + Not susceptible to number/location of bins like histograms are.
+ We can include this using the ```density()``` and ```lines()``` functions.
    + ```lines()```: takes coordinate pairs (in multiple input formats) and adds them to current figure connected by line segments
    + ```density()```: computes kernel density estimates (a smoothed histogram); see ```?density``` for more details
    
```{r echo = TRUE, eval=TRUE}
hist(prestige$prestige, freq=FALSE, col = "grey", main = "", xlab = "Prestige Score", ylim=c(0, 0.022))
abline(v = median(prestige$prestige), col = "blue", lty = 2, lwd = 2)
lines(density(prestige$prestige), col = "red", lwd = 2)
```


## Adding a Legend
+ With multiple features on one figure, a legend can help clearly convey what is plotted.
+ Use ```legend()``` to do this in R.
    + Very versatile function--look at its documentation 
    + Best to play around with this on your own, iterating through multiple plots until you get the legend to appear how you want it

```{r echo=TRUE}
hist(prestige$prestige, freq=FALSE, col = "grey", main = "", xlab = "Prestige Score", ylim=c(0, 0.022))
abline(v = median(prestige$prestige), col = "blue", lty = 2, lwd = 2)
lines(density(prestige$prestige), col = "red", lwd = 2)
legend("topright", legend = c("Median", "Density Est."),
       col = c("blue", "red"), lty = c(2, 1), lwd = 2, bty = "n")
```


## Boxplots
+ Provide a graphical representation of numeric summary statistics.
+ Use ```boxplot()``` to produce them in R
+ First let's make a boxplot of the response variable ```prestige```:
```{r echo = TRUE, eval=TRUE}
summary(prestige$prestige)
boxplot(prestige$prestige, horizontal = TRUE, xlab = "Prestige Score")
```


## Grouped Boxplots
+ Used to investigate relationships between variables:
    + Continuous variable & factor (most common)
    + Two continuous variables (group one of them)
+ Let's look at a boxplot of ```prestige``` grouped by ```type```
```{r echo = TRUE, eval=TRUE}
boxplot(prestige ~ type, data = prestige, col = "grey",
        main = "Distribution of Prestige Score by Type of Occupation",
        xlab = "Occupation Type", ylab = "Prestige Score")
```


## Scatterplots
+ We can also use scatterplots to visualize the relationship between two (continuous) variables.
+ We use the ```plot()``` function to do this
    + Extremely flexible--there are plot methods for a variety of objects! (e.g., plotting a regression object returns diagnostic plots... more on this later)
    + See the help documentation (```?plot```) and [Graphical Parameters](https://www.statmethods.net/advgraphs/parameters.html) for more details
+ For now, focus on the most simple plot where we specify the ```x``` and ```y``` coordinates.
    + ```x``` and ```y``` must be the same dimension
    
```{r echo = TRUE, eval=TRUE}
plot(x = prestige$education, y = prestige$prestige, type = "p", pch = 20,
     main = "Prestige Score by Education",
     xlab = "Avg Years of Education", ylab = "Prestige Score")
```


## Scatterplots, contd.
+ Let's overlay both a linear fit and smoother to our scatterplot. 
+ We can include this using the ```lowess()```, ```lines()```, and ```abline()``` functions.
    + We've already used ```lines()``` and ```abline()```
    + Need to use ```lm()``` to fit a linear regression, more on this later
    + ```lowess()```: computes a smoothed fit using locally-weighted polynomial regression (LOWESS); don't worry about the details here, just need to know it is an unbiased way (no assumptions) to estimate relationship between two variables
    
```{r echo = TRUE, eval=TRUE}
plot(prestige$education, prestige$prestige, type = "p", pch = 20,
     main = "Prestige Score by Education",
     xlab = "Ave. Years of Education", ylab = "Prestige Score")
abline(reg = lm(prestige ~ education, data = prestige), col = "green", lwd = 2)  # linear regression
lines(lowess(x = prestige$education, y = prestige$prestige), col = "red", lwd = 2)  # smoother
legend("topleft", legend = c("Regression Line", "Smoother"), col = c("green", "red"),
       lwd = c(2,2), lty = 1, bty = "n")
```


## Scatterplot Matrices
+ The function ```scatterplotMatrix()``` (found in the ```car``` package we installed earlier) produces scatterplots between all variables in a data frame.
+ We can use direct ordering of the variables to control the order in which they are plotted.
```{r echo = TRUE, eval=TRUE, warning=FALSE}
library(car)
scatterplotMatrix( prestige[ ,c("prestige","education","income","women")] )
```


## Next up
1. [Exercise 1](ADD ME)
2. Lunch

**Return at XXXX to discuss solutions to Exercise 2!**
