---
title: "Exercise 3 Solution"
subtitle: CSULB Intro to R
# author: Chris Galbraith, Micah Jackson
date: April 13, 2018
output: html_document
---


## Introduction
Practice subsetting the data frame you created in Exercise 2!

```{r, echo=FALSE}
name <- c("Sally", "Michelle", "Lupe", "Wendy", "Maritza",
          "Jorge", "Sam", "Joe", "Glenn", "Buck")
female <- c(TRUE, TRUE, TRUE, TRUE, TRUE, 
            FALSE, FALSE, FALSE, FALSE, FALSE)
edu <- c("HS", "BS", "BS", "MS", "PhD", 
         "HS", "HS", "BS", "MS", "PhD")
salary <- c(30, 38, 53, 65, 89, 
            29, 33, 42, 107, 246)

data <- data.frame(name = name,
                   female = female,
                   edu = edu,
                   salary = salary)

data$name <- as.character(data$name)
data$female <- as.numeric(data$female)
data$edu <- factor(data$edu, levels=c("HS", "BS", "MS", "PhD"), ordered=TRUE)
```


## Part A

Open the R script that you used for Exercise 2, and re-run the code there to ensure that ```data``` is loaded into your current environment. Check that this worked properly by typing ```data``` and ```str(data)``` at the prompt and inspecting the output.


## Part B

**B.1** Use a logical index vector to locate the rows with females. Name this vector ```femaleIndex``` and use it to look at the entries of ```data``` with females. Also, use this vector and a logical operator to look at the rows of data with men. HINT: you may find the exclamation point ```!``` useful.

```{r}
femaleIndex <- data$female == TRUE  # logical index vector selecting rows with females
data[femaleIndex, ]  # women
data[!femaleIndex, ]  # men
```


**B.2** What is the mean salary for females? Is it more or less than that of males? HINT: you may want to repeat step 1 for males, subset the column ```salary```, and use the ```mean()``` function.

```{r}
mean(data$salary[femaleIndex])  # mean salary for women
mean(data$salary[data$female == FALSE])  # mean salary for men
```

In this example, men have a higher mean salary than women.


**B.3** Is the mean salary an appropriate measure of the difference between the two groups? Compare their median salaries and comment. HINT: use the function ```median()```.

```{r}
median(data$salary[femaleIndex])  # median salary for women
median(data$salary[!femaleIndex])  # median salary for men
```

In this example, women have a higher median salary than men. The mean salary for men is skewed by Glenn and Buck's large salaries.


## Part C

**C.1** Look at the rows of ```data``` whose salaries are less than the mean salary.

```{r}
data[data$salary < mean(data$salary), ]  # entries with lower than mean salary
```


**C.2** Repeat step 1 using the median salary. Are there more or less people with salaries lower than the median versus salaries lower than the mean? Why do you think this is?

```{r}
data[data$salary < median(data$salary), ]  # entries with lower than median salary
```

In this example, there are 8 people with salaries less than the mean salary of $73,200 and 5 people with salaries less than the median of \$47,500. This is due to the skew in the data (the two abnoramlly large salaries).


## Part D

**D.1** On average, do people with a high school education make less anually than people with bachelor's degrees? Use the data to answer this question.

```{r}
mean(data$salary[data$edu == "HS"])  <  mean(data$salary[data$edu == "BS"])  # mean salary BS edu
```

Yes, in this data set people with a high school education make less than those with a bachelor's degree on average.


**D.2** Does the proportion of women differ among people with a high school education and people with at least a bachelor's degree? HINT: make a logical index vector to select people with a HS education (name it ```hs_index```) and use  the ```!``` operator to select people with at least a BS. You can use the ```sum()``` function to get the total number of ```TRUE```s from a logical index vector (e.g., the total number of poeple in the data frame with a high school education is ```sum(hs_index)```).

```{r}
hs_index <- data$edu == "HS"  # logical index vector of HS education or not
n_hs <- sum(hs_index)  # number of people in data wiht HS education
sum(data$female[hs_index]) / n_hs  # proportion of females among HS grads
sum(data$female[!hs_index]) / (nrow(data) - n_hs)  # proportion of females among college grads
```

In this example, there is a higher percentage of female college graduates than there is of highschool graduates.
