---
title: Control Structures, Functions & Packages
subtitle: CSULB Intro to R
date: April 13, 2018
#author: Chris Galbraith, Micah Jackson
output: slidy_presentation
smaller: yes
---

## Agenda
1. Control Structures
2. Functions
3. Packages
4. Helpful resources


## Control Structures

+ A control structure is a block of programming that analyzes variables and chooses a direction in which to go based on given parameters
+ Control structures in R include:
    + **for loops**
    + **if/else statements**
    + while loops 
    + repeat
    + break 
    + next
    + return
  

## for loops

+ Suppose we want to write "The year is [year]" where year is equal to 2016, 2017, and 2018. 
+ One way to do so is like this:

```{r echo = T}
print(paste("The year is", 2016))
print(paste("The year is", 2017))
print(paste("The year is", 2018))
```


## for loops

+ Or, we can use a for loop: 

```{r echo = T}
for(i in 2016:2018){
  print(paste("The year is", i))
}
```


## for loops

+ Suppose we have a numeric vector and we want to square each element
```{r echo = TRUE}
vec <- seq(2, 20, by = 2)
```
+ First, create a new vector of the same length as ```vec```
```{r echo = TRUE}
newvec <- vector("numeric", length = length(vec)) # equivalent: numeric(length(vec))
```
+ Then, write the for loop
```{r echo = TRUE}
for(i in 1:length(vec)){
  newvec[i] <- vec[i]^2
}
newvec

# advanced: vec^2
```


## if/else statements

+ if/else statements are used to write conditional statements

```{r echo = T}
x <- 7
if (x < 10){
  print("x is less than 10")
}else{
  print("x is greater than 10")
}
```

## Combining for loops and if/else statements:
+ Suppose we have the ages of 10 individuals, and we want to categorize each age as young, middle-aged, or old
+ Using a for loop, we can iterate through each age and then use if/else statements to classify each age

```{r echo = T}
age <- sample(1:100, 10)
ageCat <- rep(NA, length(age))
for (i in 1:length(age)) {
    if (age[i] <= 35){
       ageCat[i] <- "Young"
      }else if (age[i] <= 55){
        ageCat[i] <- "Middle-Aged"
      }else{
         ageCat[i] <- "Old"
      } 
}
age.df <- data.frame(age = age, ageCat = ageCat)
age.df[1:3,]
```


## Other Controls
+ break - exit the current loop or function. &nbsp;&nbsp;```break```
+ repeat - repeat a loop or function. Like an infinite loop. **Must** code in your own break condition. &nbsp;&nbsp;```repeat{do this}```
+ while loops -  condensed version of ```repeat{...; break}```. &nbsp;&nbsp; ```while(condition){do this}```
+ next - stop evaluating current iteration of your loop and move to the next. Like a minor ```break```. &nbsp;&nbsp; ```next```
+ return - like a ```break```, but exits top-level of function. Also returns an object from inside your function. &nbsp;&nbsp; ```for(){create object; return(object)}```

## Functions
+ **What:** 
    + A chunk/block of code that encapsulates a set of instructions (including other functions!).

+ **Why:** <!-- Wording ok? -->
    + Readability: Using functions makes it easier to read your code. This allows us to more quickly debug and makes it easier to make updates to our code later.
    + Reusability: Instead of rewriting a set of instructions multiple times to complete similar tasks, you can just make a call to one function.
    + Abstraction/Generalization: Functions should allow you to easily use different inputs of varying size, amount, type, etc.
    + Automation: Functions allow you to automate a task instead of manually typing out instructions
    

## How to write functions
+ Steps to writing a function
    1. Understand the purpose of the function. What do you want it to do?
    2. Define input (parameters, arguments).
    3. Determine the algorithm and output.
    4. Define data needed inside of the function (extra variables, storage, etc).
    

## Active Learning
+ What do we want our function to do?
    + Find the tallest height from a collection of heights
+ Define input.
    + "collection of heights"
    + vector of heights
    + in feet? cm? inches? miles?
    + I'm 0.00111 miles tall.
+ Determine algorithm and output.
    + output is height of tallest person (i.e. the maximum value)

    
## Active Learning
**Initialize** $tallest$ = $height_1$  
**For** each $element$ in $H$, **do**  
&nbsp;&nbsp;&nbsp;&nbsp;**If** $element > tallest$, **then**  
&nbsp;&nbsp;&nbsp;&nbsp;$tallest = element$  
end **for**  
**return** $tallest$  

```{r, evaluate=FALSE }
myHeights <- c(63, 58, 72, 65)

findTallest <- function(H){
  # initialize tallest
  tallest <- H[1]
  
  # search through all heights in H
  for(height in H){
    if(height > tallest){tallest <- height}
  }
  
  # spit out the tallest height after for-loop has run
  return(tallest)
}
```


## Up Next
+ [Exercise 4](https://datumu.github.io/CSULB_Intro_R/exercises/exercise_4/ex_4.html)
+ Packages
